Verifying Spatial Properties of Array Computations

### 2017-04-18 oopsla17/algorithms.tex
Reword lower bound condition
various edits in section 4-6

\begin{example}
\label{exm:checking}
We demonstrate analysis, checking, and inference on the 
five-point stencil example:
%%
\begin{minted}{fortran}
b(i, j) = (a(i, j) + a(i-1, j) +a (i+1, j) + a(i, j-1) + a(i, j + 1)) / 5.0
\end{minted}
\end{example}

\subsection{Analysis of array accesses}
\label{subsec:analysis}

\newcommand{\neigh}{\textsf{neigh}}
\noindent
The analysis builds on standard program analyses:
%
\textbf{(1)} basic blocks (CFG);
\textbf{(2)} induction variables per basic block;
\textbf{(3)} (interprocedural) data-flow analysis, providing a \emph{flows to}
  graph (as shorthand, the function
  $\mathit{flowTo}$ is used, implicitly parameterised by this graph,
  mapping an expression to the set of all expressions
  with forwards data-flow to this expression, based on the transitive
  closure of the flows graph);
\textbf{(4)} type information per variable.% where we use the predicate
%\arrayTy{} to classify variables of array type.

The analysis traverses the control-flow graph of a program top-down
and traverses statements inside of loops bottom-up. The right-hand
side of any assignment statement in a loop is classified based on
whether it has array subscripts flowing to it which are neighbourhood
offsets (or potential absolute in some dimensions). These are then
converted into our model domain by the interpretation
$\interp{-}^{\mathit{aterm}}$ (\Cref{defn:subscript}) and grouped
into a finite map from array variables to set of subscript schemes.
We denote sets of index/subscripts schemes by
$\mathcal{S}$.

@@

Any assignment statement from which array subscripts flow to the
current assignment is marked as visited such that the main analysis
does not classify it as the root of an array computation.  Thus,
combined with the bottom-up traversal inside of loops, we assign
specifications only to the leaves of dataflow paths in a loop.

@@

\noindent
Checking verifies the access pattern of an array computation in the
source language against any associated specifications. Checking
proceeds by generating a model from a specification and generating a
model from the source code (above), and comparing them for
consistency.  Since the model of \Cref{sec:model} interprets both
array indices and specifications as sets of points in $N$-dimensional
space. The notion of consistency is then intuitively whether these two
(potentially infinite) sets of points are equal. This leads to a
simple notion of consistency, where $\mathit{consistent}(M_C, M_S)$
tests the consistency of a modal $M_C$ of source code against
a model $M_S$ of a specification, where
$\mathit{consistent} :
    \mathsf{Mult}(\mathcal{P}(\bz{}^N)) \times
    \mathsf{Mult}(\mathsf{Approx}(\region{N})) \to \mathbb{B}$ is defined:

\begin{align*}
  \mathit{consistent}(\mathit{ix}, \mathit{model}) & = \begin{cases}
    \mathsf{false} & \mathit{model} = \mathsf{once}(x) \wedge ix =
    \mathsf{mult}(y) \\
    \mathsf{false} & \mathit{model} = \mathsf{mult}(x) \wedge ix = \mathsf{once}(y) \\
    \mathit{consistent'}(\mathit{peel}(ix), \mathit{peel}(model)) & \textit{otherwise}
  \end{cases} \\
\end{align*}
\begin{align*}
  \mathit{consistent'} & :
    \mathcal{P}(\bz{}^N) \times
    \mathsf{Approx}(\region{N}) \to \mathbb{B} & \\
%
  \mathit{consistent'}(\mathit{ix}, \mathit{model}) & = \begin{cases}
    m = \mathit{ix} & \mathit{model} = \mathsf{exact}(m) \\
    m \supseteq \mathit{ix} & \mathit{model} = \mathsf{upper}(m) \\
    m \subseteq \mathit{ix} & \mathit{model} = \mathsf{lower}(m)
  \end{cases}
\end{align*}
%
The $\mathit{consistent}$ function checks whether linearity of the
specification matches that of the indices, \ie{} if the specification allows
indices to be repeated or not. It then delegates to $\mathit{consistent'}$
to check if the points observed in the array terms match the space
defined by the specification. Lower bounds, marked \texttt{atLeast},
require the space defined by the specification to remain inside the
set of indices, while an upper bound, marked by \texttt{atMost},
requires the opposite: enclosure. In the absence of such modifiers, we expect
the space defined by observed indices corresponds exactly with those defined by the
region in the specification, hence requiring set equality.

@@

\subsection{Inferring specification automatically}
\label{subsec:inference}
%
\noindent
We provide an inference procedure for generating specifications from
code which are inserted automatically as comments. This supports
maintenance of a legacy code base, and aids adoption of the
specification language.

@@

\subsubsection{Covering}
A covering of (possibly overlapping) intervals is calculated
by coalescing \emph{contiguous} index schemes
until a fixed-point is reached.

\newcommand{\contig}[2]{\mathit{adjacent}(#1, #2)}
\begin{defn}[Adjacent]
  \label{def:contiguity}
  Two index schemes $S$ and $T$ are \emph{adjacent} written
  $\contig{S}{T}$, iff $\pi_i(S) = \pi_i(T)$ for all $1 \leq i \leq N$
  apart from some dimension $k$ such that $\pi_k(S) = [a, b]$ and
  $\pi_k(T) = [b+1, c]$.
\end{defn}
%
Given a set of indexing schemes $\mathcal{S}$ and a particular index scheme $S$
we generate a set from coalescing $S$ with adjacent index schemes:
%
\begin{equation*}
\mathit{coalesce}(S, \mathcal{S}) = 
  \{\,S \cup T \mid T \in \mathcal{S} \, \wedge \,
  \contig{S}{T} \}
\end{equation*}
%
This is then used by the following recursive procedure:
%
\begin{align*}
  \mathit{coalesceStep}(\mathcal{S}) =
  \exists T \in \mathcal{S} . 
  \begin{cases}
    \{T\} \cup \mathit{coalesceStep}(\mathcal{S} - \{T\}) &
    \mathit{coalesce}(T, \mathcal{S}) = \emptyset \\
    \mathit{coalesceStep}(\mathit{coalesce}(T, \mathcal{S}) \cup
                                        \mathcal{S} - \{T\}) & \textit{otherwise}
  \end{cases}
\end{align*}
%
This gives a specification rather than an implementation. Our
implementation represents sets by a list, and so
$\exists T \in \mathcal{S}$ above corresponds to deconstructing
the list into its head element $S$ (picking an element from the set).
If $T$ has no adjacent index schemes, then coalescing is attempted on
the rest of the set, and $T$ is returned in the result. Otherwise,
the set of coalesced index scheme is computed and this is passed to
a recursive procedure called to $\mathit{coalesceStep}$, along with the rest of
the elements).

The fixed-point of $\textit{coalesceStep}(\mathcal{S})$ is computed to
give a covering over the initial subscript space.
%
\begin{restatable}{lem}{closureInference}
  \label{lem:closuer-inference}
  For a set of index schemes $\mathcal{S}$, then
  $\mathit{coalesceSet}(\mathcal{S})$ is a set of
  index schemes in Union Normal Form.
\end{restatable}
%
For our example, the fixed-pointed of $\mathit{coalesceStep}$ is reached within two
steps:
%
\begin{align*}
  \mathit{coalesceStep}(\mathcal{S}_0) & =
  \{[-1,0] \times [0,0],\,[0,0] \times [-1, 0],\,[0,0] \times
    [0,1],\,[0,1] \times [0,0]\} \\
  \mathit{coalesceStep}^2(\mathcal{S}_0) & =
  \{[-1, 1] \times [0, 0],\,[0, 0] \times [-1, 1]\} = \mathit{coalesceStep}^3(\mathcal{S}_0)
\end{align*}
%
%\todo{Prove that this is in $Region_N$}

@@

  \item Interval schemes into a joins of meets
  of region constants, where some intervals are split into separate
  region constants.
\end{enumerate}
%
An index scheme is an interval scheme (\Cref{def:interval-scheme}) if
each vector can be represented as a vector of holed intervals
$\interv{a}{b}{c}$.  Since index schemes of the previous stage are all
contiguous and form closed intervals, a lower bound $\leq 0$ and an
upper bound $\geq 0$ is sufficient to form an interval scheme.
Otherwise, an approximate specification is generated.

An upper bound is established by
\emph{elongating} any index schemes in multiple dimensions such that
they become interval schemes. The elongation function is
given as
%
\begin{align*}
  \mathit{elongate}([a,b]) = \begin{cases}
    \interv{0}{b}{\textsf{false}} & a > 1 \\
    \interv{a}{0}{\textsf{false}} & b < -1
  \end{cases}
\end{align*}
%
If any dimension of an index scheme is elongated, then the whole index
scheme is elongated. In $\finalSet$, all index schemes that are
representable as interval schemes form a lower bound, whilst those
that do not produce an upper bound by their elongation. A lower bound
may not be generated if all indexing schemes are not interval schemes.
In the case of our example, the resulting $\finalSet$ comprises a set
of interval schemes, thus we can generate an exact specification.

@@

Recall the model of region constants in \Cref{subfig:region-model}
where $\interp{-}$ maps to holed intervals which have either $0$ as
the lower-bound or upper-bound for \fortran{forward} and
\fortran{backward} respectively, or both $0$ for \fortran{pointed}, or
$-k$ and $k$ lower and upper bounds for \fortran{centered}. We can
invert this map to generate region constants from holed
intervals. However, some intervals computed from the above steps
may not the constraints of this interpretation,
\eg{}, $\interv{-2}{1}{} \times \interv{-1}{1}{}$.  By
\Cref{lem:vector-union}, we can split interval schemes into two into
the relevant form, for example, producing
$\interv{-2}{0}{} \times \interv{-1}{1}{}$ and
$\interv{0}{1}{} \times \interv{-1}{1}{}$. 
The following function
$\textit{split}$ iterates over the components of an index scheme,
splitting apart interval schemes when necessary:
%
\begin{align*}
  %\textit{split}(\emptyset) = \emptyset \quad
  \textit{split}(\interv{a}{b}{c} \times S) =
   \begin{cases}
  \{ \interv{a}{0}{c} \times T \mid T \in \textit{split}(S) \}
  \cup
  \{ \interv{0}{b}{c} \times T \mid T \in \textit{split}(S) \}
  & a < 0 \vee b > 0 \vee (|a| \neq b) \\
  \{ \interv{a}{b}{c} \times T \mid T \in \textit{split}(S) \}
  & \textit{otherwise}
\end{cases}
\end{align*}
%
This forms a set of interval schemes to which $\interp{-}^{-1}$
(\Cref{subfig:region-model}) is well-defined for each component.

The final stage of inference, thus maps the component
of every interval scheme to a region constant via $\interp{-}^{-1}$,
combining each component with \texttt{*} can combining the
resulting regions by \texttt{+}, that is:
%
%interval schemes to region constants and combines them. %Individual
%schemes are decomposed using \Cref{lem:vector-intersect} such that
%each of the resulting index schemes have all but one dimensions set to
%\interv{-\infty}{\infty}{}.
%Each of these interval scheme are mapped to region constants using
%\Cref{subfig:region-model} and connected with \texttt{*}.
%
\begin{align}
\textit{convert}_N(\mathcal{S}) = \sum_{S \in \mathcal{S}} \; \prod_{\pi_i(S) \neq
  \interv{-\infty}{\infty}{}} \interp{\pi_i(S)}^{-1}
  \label{eq:convert}
\end{align}
%
Where summation is by the syntactic \term{+} and product is by the
syntactic \term{*}

>>>>

\begin{example}
\label{exm:checking}
We demonstrate analysis, checking, and inference on the
five-point stencil example:
%%
\begin{minted}{fortran}
b(i, j) = (a(i, j) + a(i-1, j) + a(i+1, j) + a(i, j-1) + a(i, j+1)) / 5.0
\end{minted}
\end{example}

\subsection{Static analysis of array accesses}
\label{subsec:analysis}

\newcommand{\neigh}{\textsf{neigh}}
\noindent
The analysis builds on standard program analyses:
%
\textbf{(1)} basic blocks (CFG); \textbf{(2)} induction variables per
basic block; \textbf{(3)} (interprocedural) data-flow analysis,
providing a \emph{flows to} graph (reaching definitions); and
\textbf{(4)} type information per variable.  The analysis traverses
the control-flow graph of a program top-down and traverses statements
inside of loops bottom-up. The right-hand side of any assignment
statement in a loop is classified based on whether it has array
subscripts flowing to it which are neighbourhood offsets (or a
combination of neighbourhood and absolute in some dimensions). These
are then converted into our model domain by the interpretation
$\interp{-}^{\mathit{aterm}}$ (\Cref{defn:subscript}) and grouped into
a finite map from array variables to set of subscript schemes.  We
denote sets of index/subscripts schemes by $\mathcal{S}$.

@@

Any assignment statement from which array subscripts flow to the
current assignment is marked as visited such that the main analysis
does not classify it as the root of an array computation.  In the
inference procedure, we assign specifications only to these array
computation roots.

@@

\noindent
Checking verifies the access pattern of an array computation in the
source language against any associated specifications. Checking
proceeds by generating a model from a specification and generating a
model from the source code (above), and comparing them for
consistency.  Since the model of \Cref{sec:model} interprets both
array indices and specifications as sets of points in $N$-dimensional
space, the notion of consistency is then intuitively whether these two
(potentially infinite) sets of points are equal. This leads to a
simple notion of consistency $\mathit{consistent}(M_C, M_S)$ which
tests the consistency of a model $M_C$ of source code against
a model $M_S$ of a specification, where
$\mathit{consistent} :
    \mathsf{Mult}(\mathcal{P}(\bz{}^N)) \times
    \mathsf{Mult}(\mathsf{Approx}(\region{N})) \to \mathbb{B}$ is defined:

\begin{align*}
  \mathit{consistent}(\mathit{ixs}, \mathit{model}) & = \begin{cases}
    \mathsf{false} & \mathit{model} = \mathsf{once}(x) \wedge ix =
    \mathsf{mult}(y) \\
    \mathsf{false} & \mathit{model} = \mathsf{mult}(x) \wedge ix = \mathsf{once}(y) \\
    \mathit{consistent'}(\mathit{peel}(ix), \mathit{peel}(model)) & \textit{otherwise}
  \end{cases}
\end{align*}

with the intermediate
$\mathit{consistent'} : \mathcal{P}(\bz{}^N) \times
\mathsf{Approx}(\region{N}) \to \mathbb{B}$ defined as:

\begin{align*}
  \mathit{consistent'}(\mathit{ixs}, \mathit{model}) & = \begin{cases}
    m = \mathit{ixs} & \mathit{model} = \mathsf{exact}(m) \\
    m \supseteq \mathit{ixs} & \mathit{model} = \mathsf{upper}(m) \\
    m \subseteq \mathit{ixs} & \mathit{model} = \mathsf{lower}(m)
  \end{cases}
\end{align*}
%
The $\mathit{consistent}$ function checks whether linearity of the
specification matches that of the indices, \ie{} if the specification allows
indices to be repeated or not. It then delegates to $\mathit{consistent'}$
to check if the points observed in the array terms match the space
defined by the specification. Lower bounds, marked \texttt{atLeast},
require the space defined by the specification to remain inside the
set of indices, while an upper bound, marked by \texttt{atMost},
requires the opposite: enclosure. In the absence of such modifiers, we expect
the space defined by observed indices corresponds exactly with those defined by the
region in the specification, hence requiring set equality.

@@

\subsection{Inferring specification automatically}
\label{subsec:inference}
%
\noindent
We provide an inference procedure for generating specifications from
code which are inserted automatically as comments for the root
array computations in a loop. This supports maintenance of a legacy
code base, and aids adoption of the specification language.

@@

\subsubsection{Covering}
A covering of (possibly overlapping) intervals is calculated
by coalescing \emph{adjacent} index schemes
until a fixed-point is reached.

\newcommand{\contig}[2]{\mathit{adjacent}(#1, #2)}
\begin{defn}[Adjacent]
  \label{def:contiguity}
  Two index schemes $S$ and $T$ are \emph{adjacent} written
  $\contig{S}{T}$, iff $\pi_i(S) = \pi_i(T)$ for all $1 \leq i \leq N$
  apart from some dimension $k$ such that $\pi_k(S) = [a, b]$ and
  $\pi_k(T) = [b+1, c]$.
\end{defn}
%
Given a set of indexing schemes $\mathcal{S}$ and a particular index scheme $S$
we generate a set from coalescing $S$ with adjacent index schemes:
%
\begin{equation*}
\mathit{coalesce}(S, \mathcal{S}) =
  \{\,S \cup T \mid T \in \mathcal{S} \, \wedge \,
  \contig{S}{T} \}
\end{equation*}
%
This is then used by the following recursive procedure:
%
\begin{align*}
  \mathit{coalesceStep}(\mathcal{S}) =
  \exists T \in \mathcal{S} .
  \begin{cases}
    \{T\} \cup \mathit{coalesceStep}(\mathcal{S} - \{T\}) &
    \mathit{coalesce}(T, \mathcal{S}) = \emptyset \\
    \mathit{coalesceStep}(\mathit{coalesce}(T, \mathcal{S}) \cup
                                        \mathcal{S} - \{T\}) & \textit{otherwise}
  \end{cases}
\end{align*}
%
This gives a specification rather than an implementation. Our
implementation represents sets by a list, and so
$\exists T \in \mathcal{S}$ above corresponds to deconstructing
the list into its head element $S$ (picking an element from the set).
If $T$ has no adjacent index schemes, then coalescing is attempted on
the rest of the set, and $T$ is returned in the result. Otherwise,
the set of coalesced index scheme is computed and this is passed to
a recursive procedure called $\mathit{coalesceStep}$, along with the rest of
the elements).

The fixed-point of $\textit{coalesceStep}(\mathcal{S})$ is computed to
give a covering over the initial subscript space.
%
\begin{restatable}{lem}{closureInference}
  \label{lem:closuer-inference}
  For a set of index schemes $\mathcal{S}$, then
  $\mathit{coalesceSet}(\mathcal{S})$ is a set of
  index schemes in union normal form.
\end{restatable}
%
For our example, the fixed-pointed of $\mathit{coalesceStep}$ is reached within two
steps:
%
\begin{align*}
  \mathit{coalesceStep}(\mathcal{S}_0) & =
  \{[-1,0] \times [0,0],\,[0,0] \times [-1, 0],\,[0,0] \times
    [0,1],\,[0,1] \times [0,0]\} \\
  \mathit{coalesceStep}^2(\mathcal{S}_0) & =
  \{[-1, 1] \times [0, 0],\,[0, 0] \times [-1, 1]\} = \mathit{coalesceStep}^3(\mathcal{S}_0)
\end{align*}
%
%\todo{Prove that this is in $Region_N$}

@@

  \item Interval schemes are mapped into a joins of meets
  of region constants, where some intervals split into multiple separate
  region constants.
\end{enumerate}
%
An index scheme is an interval scheme (\Cref{def:interval-scheme}) if
each vector can be represented as a vector of holed intervals
$\interv{a}{b}{c}$.  Since index schemes of the previous stage are all
adjacent and form closed intervals (in the usual mathematical sense)
then an interval with a lower bound $\leq 0$ and an
upper bound $\geq 0$ is sufficient to form an interval scheme.
Otherwise, an approximate specification is generated.

An upper bound is established by
\emph{elongating} any index schemes in multiple dimensions such that
they become interval schemes. The elongation function is
given as
%
\begin{align*}
  \mathit{elongate}([a,b]) = \begin{cases}
    \interv{0}{b}{\textsf{false}} & a > 1 \\
    \interv{a}{0}{\textsf{false}} & b < -1
  \end{cases}
\end{align*}
%
If any dimension of an index scheme is elongated, then the whole index
scheme is elongated. In $\finalSet$, all index schemes that are
representable as interval schemes form a lower bound, whilst those
that do not produce an upper bound by their elongation. A lower bound
may not be generated if none of the indexing schemes are interval schemes.
In the case of our example, the resulting $\finalSet$ comprises a set
of interval schemes, thus we can generate an exact specification.

@@

Recall the model of region constants in \Cref{subfig:region-model}
where $\interp{-}$ maps to holed intervals which have either $0$ as
the lower-bound or upper-bound for \fortran{forward} and
\fortran{backward} respectively, or both $0$ for \fortran{pointed}, or
$-k$ and $k$ lower and upper bounds for \fortran{centered}. We can
invert this map to generate region constants from holed
intervals. However, some intervals computed from the above steps
may not match the constraints of this interpretation,
\eg{}, $\interv{-2}{1}{} \times \interv{-1}{1}{}$.  By
\Cref{lem:vector-union}, we can split interval schemes into two into
the relevant form, for example, producing
$\interv{-2}{0}{} \times \interv{-1}{1}{}$ and
$\interv{0}{1}{} \times \interv{-1}{1}{}$. 
The following function
$\textit{split}$ iterates over the components of an index scheme,
splitting apart interval schemes when necessary:
%
\begin{align*}
  %\textit{split}(\emptyset) = \emptyset \quad
  \textit{split}(\interv{a}{b}{c} \times S) =
   \begin{cases}
  \{ \interv{a}{0}{c} \times T \mid T \in \textit{split}(S) \}
  \cup
  \{ \interv{0}{b}{c} \times T \mid T \in \textit{split}(S) \}
  & a < 0 \vee b > 0 \vee (|a| \neq b) \\
  \{ \interv{a}{b}{c} \times T \mid T \in \textit{split}(S) \}
  & \textit{otherwise}
\end{cases}
\end{align*}
%
This forms a set of interval schemes to which $\interp{-}^{-1}$
(\Cref{subfig:region-model}) is well-defined for each component.

The final stage of inference thus maps the component
of every interval scheme to a region constant via $\interp{-}^{-1}$,
combining each component with \texttt{*} can combining the
resulting regions by \texttt{+}, that is:
%
%interval schemes to region constants and combines them. %Individual
%schemes are decomposed using \Cref{lem:vector-intersect} such that
%each of the resulting index schemes have all but one dimensions set to
%\interv{-\infty}{\infty}{}.
%Each of these interval scheme are mapped to region constants using
%\Cref{subfig:region-model} and connected with \texttt{*}.
%
\begin{align}
\textit{convert}_N(\mathcal{S}) = \sum_{S \in \mathcal{S}} \; \prod_{\pi_i(S) \neq
  \interv{-\infty}{\infty}{}} \interp{\pi_i(S)}^{-1}
  \label{eq:convert}
\end{align}
%
Where summation is by the syntactic \term{+} and product is by the
syntactic \term{*}.
